### 2014-03-25 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
figure
mer

### Overview
This report aims to show to what extent the use of ComBat led to false results in the second analysis example given in [Johnson et al.](http://biostatistics.oxfordjournals.org/content/8/1/118.abstract) The example named "Data Set 2" and the analysis are described in the [supplementary material](http://biostatistics.oxfordjournals.org/content/suppl/2006/04/21/kxj037.DC1/kxj037supp.pdf) for Johnson et al.
Description on how to obtain the data were found [here](https://groups.google.com/forum/#!msg/combat-user-forum/S9vBcXw8RGk/QIoD6oRBM9IJ) which had links to [dataExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/dataExample2.txt)
and [sampleInfoExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/sampleInfoExample2.txt)

This document has four main parts
- Reproduce some of the results to show that we are working on the same data and analysis workflow
- Remove the use of ComBat and perform the same analysis with an alternative established tool
- Estimate the error introduced by ComBat and the consequences for the conclusion of the study
- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is mainly due to error introduced by ComBat

@@

```
Inspection of the covariate/batch balance;
```{r}
print(table(sampleannotation[ , c("Batch", "Type")])  )
```

@@

We start with the data before ComBat adjustment. Limma works best with log transformed and between array normalized values, so first we set negative values to 1 and filter out probes that has this in more than half the samples. To ease later comparisons the same filter will also be applied to the EB3 data set. Then the test is run with the batch included as a blocking factor.
```{r}
negativeprobesfilter =( rowSums(datamatrix>=1) >= (0.5*ncol(datamatrix)) ) & 
                      ( rowSums(EB3>=1) >= (0.5*ncol(EB3)) )
thismat = datamatrix[negativeprobesfilter, ]
thismat[thismat<1]=1
thismat=normalizeBetweenArrays(log2(thismat))
Type = as.factor(sampleannotation$Type)
Block = as.factor(sampleannotation$Batch)
design = model.matrix(~0+Type+Block)
fit = lmFit(thismat, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)	
fit2 = contrasts.fit(fit, cont.matrix)
datamatrix_limma_res = eBayes(fit2)
rm(thismat, Type, Block, design, fit, cont.matrix, fit2)
print(table(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05))
```
Number of differentially expressed probes found with the alternative anaysis is **`r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)`**.

@@

Lets assume one aim of this study was to obtain a list of genes that differed between control and treatment with a significant cut-off of q<0.05. Then handling the batch effect with ComBat adjustment will yield a list of **`r sum(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05)`** whereas a list of **`r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)`** probes would have been closer to the truth. Also it is of interest to note that it is mostly the same probes that are found.
```{r, tidy=FALSE}
table( qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05 ,
         qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05,
       dnn=c("limma", "ComBat"))
```
And if they were to use the top 1000 probes in a gene set test, they would have found many of the same genes regardless of hadling batch effects by ComBat or limma.
```{r, tidy=FALSE}
table( rank(datamatrix_limma_res$p.value[,1])  <=1000 & 
         rank(EB3_limma_res$p.value[,1])<=1000   )
```
Our conclusion is that for this study the error introduced by the use of ComBat would probably have a modest effect on the final result.

@@

```
Plotting the p-values
```{r, dev='svg', fig.width=8, fig.height=8}
hist(EBrand_limma_pvalues, border="blue",  main="P-values, Random numbers", breaks=100, xlab="p-value")
hist(randommatrix_limma_pvalues, border="black", add=T, breaks=100)
hist(randommatrix_limma_batch_pvalues, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted", "Limma adjusted", "No adjustment"), text.col=c("blue", "black", "red"))
```
The p-values for ComBat adjusted random numbers are slightly enriched for low p-values. Indicating that the same enrichment seen in the real data is also false.

@@

```{r, tidy=FALSE}
Batch45 = Batch3
Batch45_annot = sampleannotation[dimnames(Batch45)[[2]],]
Batch45_annot$Batch="4"
Batch45_annot$Batch[Batch45_annot$Type=="C"][1:4] = "5"
Batch45_annot$Batch[Batch45_annot$Type=="R"][1:3] = "5"
print(table(sampleannotation[sampleannotation$Batch %in% c("1", "2"), c("Batch", "Type")]))
print(table(Batch45_annot[, c("Batch", "Type")]))
```
Then adjust the data with fictive batches and calculate p-values as done for the unadjusted batch 3.
```{r, results='hide', tidy=FALSE}
EB45 = as.matrix( sva::ComBat(
            dat=Batch45, 
            batch=Batch45_annot[colnames(Batch45),"Batch"], 
            mod=model.matrix( ~as.factor(Batch45_annot[colnames(Batch45),"Type"])  ), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))

>>>>

### Overview
This report aims to show to what extent the use of ComBat led to false results in the second analysis example given in [Johnson et al.](http://biostatistics.oxfordjournals.org/content/8/1/118.abstract) The example named "Data Set 2" and the analysis is described in the [supplementary material](http://biostatistics.oxfordjournals.org/content/suppl/2006/04/21/kxj037.DC1/kxj037supp.pdf) for Johnson et al.
Description on how to obtain the data were found [here](https://groups.google.com/forum/#!msg/combat-user-forum/S9vBcXw8RGk/QIoD6oRBM9IJ) which had links to [dataExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/dataExample2.txt)
and [sampleInfoExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/sampleInfoExample2.txt)

This document has four main parts 
- Reproduce some of the results to show that we are working on the same data and analysis workflow
- Remove the use of ComBat and perform the same analysis with an alternative established tool
- Estimate the error introduced by ComBat and the consequences for the conclusion of the study
- Perform a few more sanity checks to substantiate that the difference in results for the two above analyses is mainly false and introduced by ComBat

@@

```
Inspection of the covariate/batch balance;
```{r}
print(table(sampleannotation[ , c("Batch", "Type")])  )
```
Three batches and two covariates. There is a "Cell" column in the sample annotation, but it does not seem to have been used.

@@

We start with the data before ComBat adjustment. Limma works best with log transformed and between array normalized values, so first we set negative values to 1 and filter out probes that has this in more than half the samples. To ease later comparisons the same filter will also be applied to the EB3 data set. Then the test is run with the batch included as a blocking factor.
```{r}
negativeprobesfilter =( rowSums(datamatrix>=1) >= (0.5*ncol(datamatrix)) ) & 
                      ( rowSums(EB3>=1) >= (0.5*ncol(EB3)) )
thismat = datamatrix[negativeprobesfilter, ]
thismat[thismat<1]=1
thismat=normalizeBetweenArrays(log2(thismat))
Type = as.factor(sampleannotation$Type)
Block = as.factor(sampleannotation$Batch)
design = model.matrix(~0+Type+Block)
fit = lmFit(thismat, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)	
fit2 = contrasts.fit(fit, cont.matrix)
datamatrix_limma_res = eBayes(fit2)
rm(thismat, Type, Block, design, fit, cont.matrix, fit2)
print(table(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05))
```
Number of differentially expressed probes found with the alternative analysis is **`r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)`**.

@@

Lets assume one aim of this study was to obtain a list of genes that differed between control and treatment with a significant cut-off of q<0.05. Then handling the batch effect with ComBat adjustment will yield a list of **`r sum(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05)`** whereas a list of **`r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)`** probes would have been closer to the truth. Also it is of interest to note that it is mostly the same probes that are found.
```{r, tidy=FALSE}
table( qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05 ,
         qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05,
       dnn=c("limma", "ComBat"))
```
And if they were to use the top 1000 probes in a gene set test, they would have found many of the same genes regardless of handling batch effects by ComBat or limma.
```{r, tidy=FALSE}
table( rank(datamatrix_limma_res$p.value[,1])  <=1000 & 
         rank(EB3_limma_res$p.value[,1])<=1000   )
```
Our conclusion is that for this study the error introduced by the use of ComBat would probably have a modest effect on the final result.

@@

```
Plotting the p-values
```{r, dev='svg', fig.width=8, fig.height=8}
hist(EBrand_limma_pvalues, border="blue",  main="P-values, Random numbers", breaks=100, xlab="p-value")
hist(randommatrix_limma_pvalues, border="black", add=T, breaks=100)
hist(randommatrix_limma_batch_pvalues, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted", "Limma adjusted", "No adjustment"), text.col=c("blue", "black", "red"))
```
The p-values for ComBat adjusted random numbers are slightly enriched for low p-values. Indicating that some of the enrichment seen in the real data is also false.

@@

```{r, tidy=FALSE}
Batch45 = Batch3
Batch45_annot = sampleannotation[dimnames(Batch45)[[2]],]
Batch45_annot$Batch="4"
Batch45_annot$Batch[Batch45_annot$Type=="C"][1:4] = "5"
Batch45_annot$Batch[Batch45_annot$Type=="R"][1:3] = "5"
print(table(sampleannotation[sampleannotation$Batch %in% c("1", "2"), c("Batch", "Type")]))
print(table(Batch45_annot[, c("Batch", "Type")]))
```
Then adjust the data with fictive batches and calculate p-values as done for the un-adjusted batch 3.
```{r, results='hide', tidy=FALSE}
EB45 = as.matrix( sva::ComBat(
            dat=Batch45, 
            batch=Batch45_annot[colnames(Batch45),"Batch"], 
            mod=model.matrix( ~as.factor(Batch45_annot[colnames(Batch45),"Type"])  ), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
